
---
title: "Statistical inference: MCMC methods"
author: "Frank van der Meulen (TU Delft)"
date: ""
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r}
setwd("~/.julia/dev/WI4455/scr/mcmc")
library(tidyverse)
library(gridExtra)
```

# SIMPLE EXAMPLE OF METROPOLIS-HASTINGS ALGORITHM

Suppose we wish to simulate from the beta distribution.  We can do this with the Metropolis–Hastings algorithm, where the target density is the $\beta(a,b)$-distribution. Of course this is just an example for illustration, as there exist direct ways for simulating independent realisations of the beta distribution.

## MH with independent $Unif(0,1)$-proposals. 
```{r}
  a=2.7; b=6.3; # choose parameters of the beta-distribution
  
  Nsim = 5000
  X = c(runif(1), rep(0,Nsim-1))  # initialize the chain
  acc = rep(0,Nsim)
  for (i in 2:Nsim)
  {
    Y=runif(1)  # proposal
    A=dbeta(Y,a,b)/dbeta(X[i-1],a,b)
    acc[i-1] <- (runif(1)<A) 
    X[i]=X[i-1] + (Y-X[i-1])*acc[i-1]
  }
  df = data.frame(iterate=1:Nsim, vals=X)
  p1 = df %>% ggplot(aes(x=iterate, y=vals)) + geom_line()  + ylab("")
  p2 = ggplot()  + geom_histogram(data=df, mapping=aes(x=vals,y=..density..), colour='white', bins=50) +  
      xlab("") + ylab("") +
      stat_function(data=data.frame(x=c(0,1)), mapping=aes(x), fun=function(x) dbeta(x,a,b),colour="orange", size=1.2)
    
  
pdf('Beta-ind.pdf',width=7,height=4)
grid.arrange(p1,p2)
dev.off()
cat('average acceptance probability equals: ',mean(acc))
```

## MH with smmetric random-walk MH proposals. 

If $x$ is the current iterate, then propose $x+U(-\eta,\eta)$. Experiment yourself with different values for $\eta$. 
```{r}
  eta=10
  X=rep(runif(1),Nsim)  # initialize the chain
  acc=rep(0,Nsim)
  for (i in 2:Nsim)
  {
    Y=X[i-1] + runif(1,-eta,eta)
    A=dbeta(Y,a,b)/dbeta(X[i-1],a,b)
    acc[i-1] <- (runif(1)<A) 
    X[i]=X[i-1] + (Y-X[i-1])*acc[i-1]
  }


df = data.frame(iterate=1:Nsim, vals=X)
p1 = df %>% ggplot(aes(x=iterate, y=vals)) + geom_line()  + ylab("")
p2 = ggplot()  + geom_histogram(data=df, mapping=aes(x=vals,y=..density..), colour='white', bins=50) +  
      xlab("") + ylab("") +
      stat_function(data=data.frame(x=c(0,1)), mapping=aes(x), fun=function(x) dbeta(x,a,b),colour="orange", size=1.2)
    
  
pdf('Beta-rw10.pdf',width=7,height=4)  
grid.arrange(p1,p2)
dev.off()
cat('average acceptance probability equals: ',mean(acc))
```

# BASEBALL EXAMPLE, MODEL 2

The data are
```{r}
players <- c('McGwire', 'Sosa', 'Griffey', 'Castilla', 'Gonzalez', 'Galaragga',  'Palmeiro', 
'Vaughn',' Bonds', 'Bagwell', 'Piazza', 'Thome', 'Thomas','T. Martinez', 'Walker', 'Burks', 'Buhner')
Y <- c(7,9,4,7,3,6,2,10,2,2,4,3,2,5,3,2,6)
n <- c(58,59,74,84,69,63,60,54,53,60,66,66,72,64,42,38,58)
AB <- c(509,643,633,645,606,555,619,609,552,540,561,440,585,531,454,504,244)
HR <- c(70,66,56,46,45,44,43,40,37,34,32,30,29,28,23,21,15)
print(baseball<-data.frame(players=players,PS_AB=n,PS_HR=Y,S_AB=AB,S_HR=HR))
```


We need the following 2 functions for updating the coefficients $\mu_i$, $i=1,\ldots, 17$. 
```{r}
logtargetMui <- function(Yi, ni, mui, th, tau2,tunePar)
    Yi*mui-((mui-th)^2)/(2*tau2)-ni*log(1+exp(mui))

updateMui <- function(Yi, ni, mui, th, tau2, tunePar)
{
  muiNew <- mui + tunePar * rnorm(1)
  A <- exp(logtargetMui(Yi,ni,muiNew,th, tau2,tunePar)-
             logtargetMui(Yi,ni,mui,th, tau2,tunePar))
  ifelse (runif(1)<A, muiNew, mui)
}
```

```{r}
N <- length(Y)
IT <- 10000  # number of iterations
tunePar <- 1 # tuning par for MH step updating mu (sd of normal distr)

# prior hyperpars
alpha <- 0.001
beta <- 0.001

# save iterates in matrix and vectors
mu <- matrix(0,IT,N)
th <- rep(0,IT)
tau2 <- rep(0,IT)

# initialise
mu[1,] <- rnorm(N,sd=5)  # arbitrary 
th[1] <- rnorm(1)
tau2[1] <-2

# Gibbs sampler:
for (it in 2:IT)
{
  # update mu
  for (i in 1:N)
  {  mu[it,i] <- updateMui(Y[i],n[i],mu[it-1,i],
                           th[it-1],tau2[it-1],tunePar)  }
  # update th
  th[it] <- rnorm(1,mean(mu[it,]),sqrt(tau2[it-1]/N))
  # update tau2
  tau2[it] <- 1/(rgamma(1,shape=N/2+alpha, 
                       rate=beta+.5*sum((mu[it,]-th[it])^2)))
}  
```

Visualisation
```{r}
par(mfrow=c(3,2)) # make trace plots for th, tau2, mu1, mu2, mu3, mu4
plot.ts(th);plot.ts(tau2)
plot.ts(mu[,1]);plot.ts(mu[,2])
plot.ts(mu[,3]);plot.ts(mu[,4])

BI <- 1000 # discard first BI samples as BurnIn

plot.ts(th[BI:IT]);plot.ts(tau2[BI:IT])
plot.ts(mu[BI:IT,1]);plot.ts(mu[BI:IT,2])
plot.ts(mu[BI:IT,3]);plot.ts(mu[BI:IT,4])

# computate posterior means
th.pm <- mean(th[BI:IT])
tau2.pm <- mean(tau2[BI:IT])
mu.pm <- colMeans(mu[BI:IT,])
p.pm <- colMeans(1/(1+exp(-mu[BI:IT,])))

print(th.pm)
print(tau2.pm)
print(mu.pm)
```

Add results to baseball dataframe
```{r}
baseball$bayes <- p.pm
baseball$mle <- baseball$PS_HR/baseball$PS_AB # equals empirical fraction
baseball
library(ggplot2)
p1 <- ggplot(baseball, aes(x=players)) + 
  geom_point(aes(y = bayes, shape="Bayes"),size=2.5) + 
  geom_point(aes(y = mle, shape="mle"),size=2.5)+
  theme_minimal()+
    theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5),legend.position='none') +ylab('probability') +ggtitle("Pre-season data")+xlab("")
```

Add predictions for season
```{r}
baseball$S_HR_Bayes <- baseball$S_AB * baseball$bayes
baseball$S_HR_mle <- baseball$S_AB * baseball$mle
baseball
```

Compare performance
```{r}
performance_Bayes = sum((baseball$S_HR-baseball$S_HR_Bayes)^2)
performance_mle = sum((baseball$S_HR-baseball$S_HR_mle)^2)
p2<- ggplot(baseball, aes(x=players)) + 
     geom_point(aes(y = S_HR_Bayes, shape ="Bayes"),size=2.5)+
  geom_point(aes(y = S_HR_mle, shape = "mle"),size=2.5)+
  geom_point(aes(y = S_HR),shape=8,size=2.5,colour='blue')+theme_light()+
    theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5),legend.position='bottom') +ylab('nr of homeruns') +ggtitle("Comparison predicted (using pre-season data) and observed (seasan data)")
pdf('output-baseball_combined.pdf',width=8,height=8)
grid.arrange(p1,p2,ncol=1)
dev.off()

cat('performance Bayes equals: ',performance_Bayes)
cat('performance mle equals: ',performance_mle)
```

Compute predictive distributions
```{r}
ind <- seq(BI,IT,by=10) # use every 10-th iterate from non-burnin samples
L <- length(ind)
pred <- matrix(0,L,N)

for (i in 1:N)
{
    for (j in 1:L)
        pred[j,i] <- rbinom(1,AB[i], 1/(1+exp(-mu[ind[j],i])))
}

meanPred <- colMeans(pred)
meanPred
cat('Sum of squared prediction error equals: ',sum((meanPred-HR)^2))

# plot for the i-th second player the predictive distr
i<-16
hist(pred[,i],breaks='FD',prob=TRUE, main='predictive distribution',xlab='',col='lightblue')
lines(density(pred[,i]),lwd=2.5)
```

Note that the results are sensitive to the choice of hyperpars. $\alpha=\beta=0.01$ and $\alpha=\beta=0.001$ give quite different sum of squared prediction errors! 